\textbf{The Structure of Chaos: An Empirical Comparison of Fractal Physiology Complexity Indices using NeuroKit2}

### 2022-07-26 analysis.Rmd
format

Complexity is an umbrella term for concepts derived from information theory, chaos theory, and fractal mathematics, used to quantify unpredictability, entropy, and/or randomness. Using these approaches to characterize physiological signals [a subfield commonly referred to as "fractal physiology", @bassingthwaighte2013fractal] has shown promising results in the assessment and diagnostic of the state and health of living systems [@lau2021brain; @ehlers1995chaos; @goetz2007].

@@

Additionally, the lack of a comprehensive open-source and user-friendly software for computing various complexity indices likely contributes to the scarcity of empirical comparisons [@flood2021]. Indeed, many complexity indices are only described mathematically in journal articles, with reusable code seldom made available, therefore limiting their further application and validation [@yang2013; @flood2021]. Even when available and open-source, the code implementations of complexity measures are typically found scattered across different packages or scripts, or embedded within a larger goal-directed framework [e.g., *HCTSA*, a time-series comparison tool, @fulcher2017]. To address this lack of unified accessibility, we added a comprehensive set of complexity-related features to *NeuroKit2*, a Python package for physiological signal processing [@Makowski2021neurokit]. In doing so, we hope to provide users with an easy-to-use software capable of computing a wide range of complexity measures. The code is designed to run quickly and efficiently, while still being written in pure Python [with the help of standard dependencies such as *NumPy* or *Pandas*, @harris2020array; @mckinney2010data] to maximize its reusability, transparency, and correctness.

@@

We started by generating 6 types of signals, one random-walk, two oscillatory signals (with one made of harmonic frequencies that results in a self-repeating - fractal-like - signal), two complex signals derived from Lorenz systems (with parameters [$\sigma = 10, \beta = 2.5, \rho = 28$]; and [$\sigma = 20, \beta = 2, \rho = 30$], respectively) and one EEG-like simulated signal. Each of these signals was iteratively generated at 6 different lengths (ranging from 500 to 3000 by 500 samples). The resulting vectors were standardized and each were added 5 types of $(1/f)^\beta$ noise (namely violet $\beta=-2$, blue $\beta=-1$, white $\beta=0$, pink $\beta=1$, and brown $\beta=2$ noise). Each noise type was added at 128 different intensities (linearly ranging from 0.001 to 3). Examples of generated signals are presented in **Figure 1**.

@@

As the span and application of complexity science grows, a systematic approach to compare their "performance" becomes necessary to reinforce the clarity and structure of the field. The term *performance* used here is to be understood in a relative sense, as any such endeavor faces the "hard problem" of complexity science: various objective properties of signals [e.g., short-term vs. long-term variability, auto-correlation, information, randomness, @namdari2019; @xiong2017] participate in forming overarching concepts such as "complex" and "chaotic". Indices that are sensitive to some of these objective properties are thus conceptually linked through such overarching frameworks. However, it remains unclear how these high-level concepts transfer back, in a top-down fashion, into a combination of lower-level features. As such, it is conceptually complicated to benchmark complexity measures against "objectively" complex vs. non-complex signals. In other words, while we are aware that different objective signal characteristics can contribute to the "complexity" of a signal, there is not a one-to-one correspondence between the latter and the former.

>>>>

Complexity is an umbrella term for concepts derived from information theory, chaos theory, and fractal mathematics, used to quantify unpredictability, entropy, and/or randomness. Using these approaches to characterize physiological signals (a subfield commonly referred to as "fractal physiology" [@bassingthwaighte2013fractal]) has shown promising results in the assessment and diagnostic of the state and health of living systems [@lau2021brain; @ehlers1995chaos; @goetz2007].

@@

Additionally, the lack of a comprehensive open-source and user-friendly software for computing various complexity indices likely contributes to the scarcity of empirical comparisons [@flood2021]. Indeed, many complexity indices are only described mathematically in journal articles, with reusable code seldom made available, therefore limiting their further application and validation [@yang2013; @flood2021]. Even when available and open-source, the code implementations of complexity measures are typically found scattered across different packages or scripts, or embedded within a larger goal-directed framework (e.g., *HCTSA*, a time-series comparison tool [@fulcher2017]). To address this lack of unified accessibility, we added a comprehensive set of complexity-related features to *NeuroKit2*, a Python package for physiological signal processing [@Makowski2021neurokit]. In doing so, we hope to provide users with an easy-to-use software capable of computing a wide range of complexity measures. The code is designed to run quickly and efficiently, while still being written in pure Python (with the help of standard dependencies such as *NumPy* or *Pandas* [@harris2020array; @mckinney2010data]) to maximize its reusability, transparency, and correctness.

@@

We started by generating 6 types of signals, one random-walk, two oscillatory signals (with one made of harmonic frequencies that results in a self-repeating - fractal-like - signal), two complex signals derived from Lorenz systems (with parameters $\sigma = 10, \beta = 2.5, \rho = 28$; and $\sigma = 20, \beta = 2, \rho = 30$, respectively) and one EEG-like simulated signal. Each of these signals was iteratively generated at 6 different lengths (ranging from 500 to 3000 by 500 samples). The resulting vectors were standardized and each were added 5 types of $(1/f)^\beta$ noise (namely violet $\beta=-2$, blue $\beta=-1$, white $\beta=0$, pink $\beta=1$, and brown $\beta=2$ noise). Each noise type was added at 128 different intensities (linearly ranging from 0.001 to 3). Examples of generated signals are presented in **Figure 1**.

@@

As the span and application of complexity science grows, a systematic approach to compare their "performance" becomes necessary to reinforce the clarity and structure of the field. The term *performance* used here is to be understood in a relative sense, as any such endeavor faces the "hard problem" of complexity science: various objective properties of signals (e.g., short-term vs. long-term variability, auto-correlation, information, randomness [@namdari2019; @xiong2017]) participate in forming overarching concepts such as "complex" and "chaotic". Indices that are sensitive to some of these objective properties are thus conceptually linked through such overarching frameworks. However, it remains unclear how these high-level concepts transfer back, in a top-down fashion, into a combination of lower-level features. As such, it is conceptually complicated to benchmark complexity measures against "objectively" complex vs. non-complex signals. In other words, while we are aware that different objective signal characteristics can contribute to the "complexity" of a signal, there is not a one-to-one correspondence between the latter and the former.
